---
title: "Geographic cleaning of occurrence data from errors common to biological databases"
author: "Alex Zizka"
date: "16 September 2017"
output: html_document
---

## Background
Issues with data quality are a central issue hampering the use of publicly available species occurrence data in ecology and biogeography. Major problems are: data entry errors leading to erroneous occurrence records, imprecise geo-referencing mostly of pre-GPS records and missing metadata specifying data entry and coordinate precision. Manual data cleaning based on expert knowledge can mostly detect these issues, but is only applicable for small taxonomic or geographic scales and is difficult to reproduce. Automated clean procedures are more scalable alternative.

## Outcomes
After this exercise you will be able to:
1. Identify common problems in coordinates of biological collection datasets
2. Use GBIF provided meta-data to improve coordinate quality in R
3. Use automated cleaning algorithms to flag common problems

## Exercise
In this exercise we will use the lion example data ("bombacoideae_occurrences_gbif.txt"), alternatively you can use your own data, or data obtained during the first exercise. You can find potentially useful R functions for each question in the brackets. Get help for all functions with ?FUNCTIONNAME. If you have no experience with R, but want to get familiar with the cleaning functions of speciesgeocodeR, please check out the graphical-user-interface app at https://azizka.shinyapps.io/CoordinateCleaner/. You can also find more tutorials at the [speciesgeocodeR wiki](https://github.com/azizka/speciesgeocodeR/wiki).

1. Load the example data (we will use "bombacoideae_occurrences_gbif.csv"), and limit the data to columns with potentially useful information (`read_csv`, `names`, `select`).
2. Visualize the coordinates on a map (`borders`, `ggplot`, `geom_point`).
3. Clean the coordinates based on available meta-data. A good starting point is to plot continuous variables as histogram and look at the values for discrete variables. Remove unsuitable records and plot again (`table`, `filter`).
4. Apply the automated cleaning as implemented in speciesgeocodeR, to remove problematic records (`CleanCoordinates`, `plot`).
5. Apply the automated dataset level cleaning as implemented in speciesgeocodeR to see if there are potential problems with coordinate precision (`CleanCoordinatesDS`). 


## Library setup
You will need the following R libraries for this exercise, just copy the code chunk into you R console to load them. You might need to install some of them separately.

```
require(speciesgeocodeR)
require(tidyverse)
require(countrycode)

```
# Solutions
The following code is a suggestion for data cleaning and are by no means guaranteed to be exhaustive or a one-size-fits it all solution. The hope is that they will significantly improve the dataset compared to the raw data. *Remember:* What  is 'good data' depends completely on the type of downstream analyses and their spatial scale. The cleaning here might be a good starting point for continental scale macroecological analyses or ancestral area estimation in historical biogeography. For this exercise we will simply exclude potentially problematic records, but it is of course also possible to go back into GBIF and double check each record thoroughly.

## 1. Load the example data, and visualize the coordinates on a map.

GBIF provides a large amount of information for each record, leading to a huge data.frame with many columns. However some of this information is only available for few records, and thus for most analyses most of the columns can be dropped. Here, we will only retain information to identify the record and information that is important for cleaning up the data.

```{r}
dat <- read_csv("Example_data/bombacoideae_occurrences_gbif.csv")

names(dat) #a lot of columns

dat <- dat %>%
  select(species, decimalLongitude, decimalLatitude, countryCode, individualCount,
         gbifID, family, taxonRank, coordinateUncertaintyInMeters, year,
         basisOfRecord, institutionCode, datasetName) # you might find other ones useful depending on your downstream analyses

```


## 2. Visualize the coordinates on a map
Visualizing the data on a map can be extremely helpful to understand potential problems and to identify problematic records. 

```{r}
wm <- borders("world", colour="gray50", fill="gray50")
ggplot() + 
  coord_fixed()+
  wm + 
  geom_point(data = dat, aes(x = decimalLongitude, y = decimalLatitude), 
             colour = "darkblue", size = 0.5)
```


## 3. Clean the coordinates based on available meta-data
As you cans see there are a lot of unexpected occurrence locations, outside the current distribution range. We will find out the reasons for this in a minute. In this specific case we could relatively easily get rid of a large number of problematic records based on prior knowledge (lions should not occur outside Africa and India) but we usually do not have this kind of knowledge when dealing with larger datasets, so we will try to get rid of those records in different ways. GBIF data often contain a good number of meta-data that can help to locate problems. First we'll remove data without coordinates, coordinates with very low precision and the unsuitable data sources. WE will remove all records with a precision below 100km as this represent the grain size of many macro-ecological analyses, but the number is somewhat arbitrary and you best chose it based on your downstream analyses. We also exclude fossils as we are interested in recent distributions and records of unknown source, as we might deem them not reliable enough.

```{r}
#remove records without coordinates
dat.cl <- filter(dat, !is.na(decimalLongitude) & !is.na(decimalLatitude))

#remove records with low coordinate precision
hist(dat.cl$coordinateUncertaintyInMeters/1000)

dat.cl <- dat.cl %>% 
  filter(coordinateUncertaintyInMeters/1000 <= 100 | is.na(coordinateUncertaintyInMeters))

#remove unsuitable data sources, especially fossils
table(dat$basisOfRecord)

dat.cl <- filter(dat.cl, basisOfRecord == "HUMAN_OBSERVATION" | basisOfRecord == "OBSERVATION" |
                         basisOfRecord == "PRESERVED_SPECIMEN")
```


In the next step we will remove records with suspicious individual counts. GBIF for some reason also includes records of absence and suspiciously high occurrence counts might indicated inappropriate data or data entry problems. We might also want to exclude very old records, as they might be unreliable. Records from before the second world war are often unreliable, if they were georeferenced based on political entities, additionally old records might be likely form areas where species went extinct (for example due to land-use change).

```{r}
#Individual count
table(dat.cl$individualCount)

dat.cl <- dat.cl%>%
  filter(individualCount > 0 | is.na(individualCount))%>%
  filter(individualCount < 99 | is.na(individualCount)) # high counts are not a problem here

#age of record
table(dat.cl$year)

dat.cl <- dat.cl%>%
  filter(year > 1945)
```

On top of the geographic cleaning, we also want to make sure to only include species level records and records from the right taxon. The latter is not a problem in this case, as we only have one species, but it can be helpful for large datasets. Taxonomic problems such as spelling mistakes in the names or synonyms can be a severe problem. We'll not treat taxonomic cleaning here, but check out the [taxize R package](https://ropensci.org/tutorials/taxize_tutorial.html) or the [taxonomic name resolution service](http://tnrs.iplantcollaborative.org/) for that.

```{r}
table(dat.cl$family) #that looks good
dat.cl <- dat.cl%>%
  filter(family == 'Malvaceae')

table(dat.cl$taxonRank) # this is also good

```

We excluded almost 70% of the initial data points with the data cleaning! Most of them due to missing coordinates, but the general picture has improved considerably as well. We plot again.

```{r}
wm <- borders("world", colour="gray50", fill="gray50")
ggplot() + 
  coord_fixed()+
  wm + 
  geom_point(data = dat.cl, aes(x = decimalLongitude, y = decimalLatitude), 
             colour = "darkblue", size = 0.5)

```

## 4. Apply the automated cleaning as implemented in speciesgeocodeR
As a first step we will run the automatic cleaning algorithm of speciesgeocodeR. The `CleanCoordinates` function implements a large set of automated cleaning steps to flag errors that are common to coordinates from biological collections, including among others: sea coordinates, zero coordinates, coordinate - country mismatches, coordinates assigned to country and province centroids, outlier coordinates and coordinates assigned to biodiversity institutions. You can switch on each test individually and provide custom gazetteers for each test. To use the country - coordinate mismatch test we need to convert the country from ISO2 to ISO3 format. 

```{r}
#convert country code
dat.cl$countryCode <-  countrycode(dat.cl$countryCode, origin =  'iso2c', destination = 'iso3c')

#flag problems
flags <- CleanCoordinates(x = dat.cl[,2:3], countries = dat.cl$countryCode, species = dat.cl$species, countrycheck = T, outliers = T)

#Exclude problematic records
dat.cl <- dat.cl[flags$summary,]

```

So we were able to additionally remove a substantial number of records with the automated cleaning. However, be aware that some potential problems remained, such as the records in Europe and northern North America, which most likely represent cultivated individuals we could not exclude automatically.

## 5. Apply the automated dataset level cleaning as implemented in speciesgeocodeR
Some problems, in particular certain kinds of imprecisions, cannot be identified on the record level. For instance, many records are based on gridded sampling schemes or atlas projects, but are not easily identifiable as such. To identify these kind of problems speciesgeocodeR includes also dataset level tests, which search for periodicity in the decimals of occurrence records, and can indicate, if a substantial portion of the coordinates in a dataset have been subject to rounding or are nodes of a raster scheme. You can run this test either on the entire dataset, or on individual contributing dataset, e.g. all records from one museum, using the `CleanCoordinatesDS` function. See [here](https://github.com/azizka/speciesgeocodeR/wiki/1.-Automated-Cleaning-of-Geographic-Data) for more details.
```{r}
#For the total dataset
dat.cl$datasettotal <- "TOTAL"

t.all <- dat.cl%>%
  select(dataset = datasettotal, 
         decimallongitude = decimalLongitude, 
         decimallatitude = decimalLatitude)

##Run dataset level test
CleanCoordinatesDS(t.all)

#For individual datasets, this will take some time
##Remove small datasets
sel <- table(dat.cl$datasetName)
sel <- sel[sel > 500]

t.ind <- dat.cl%>%
  filter(dat.cl$datasetName %in% names(sel))%>%
  select(dataset = datasetName, 
         decimallongitude = decimalLongitude, 
         decimallatitude = decimalLatitude)

##Run dataset level test
CleanCoordinatesDS(t.ind)
```

There is no evidence for periodicity in the entire dataset or its three biggest contributing datasets. Great!
